Inference of Bifurcations\\with Differentiable Continuation

### 2021-05-17 docs/article.tex

For clarity, we guide the reader through the methods with the following minimal models that explore the space of saddle-nodes $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchforks $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$. These minimal models are shown with targets $\targets$ in Figure \ref{fig:saddle-node} and Figure \ref{fig:pitchfork} respectively. These figures show that the determinant $\Det=0$ at the bifurcations and that its directional derivative along the bifurcation curve $\frac{d}{ds}\Det$ is finite and non-zero, at least in the non-degenerate case. This gives us our first hint of what should be optimised to increase the number of predictions $|\predictions(\theta)|$.

\begin{figure}
\centering
\includegraphics[width=6cm]{saddle-node}
\includegraphics[width=6cm]{pitchfork}
\caption{Saddle-node model $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchfork model $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$ with $\theta=(5/2,-1),(1/2,-1)$ and targets $\targets=\{-1,1\},\{0\}$ respectively. Determinant in red. Lighter shades indicate the determinant crossing zero for unstable solutions}
\label{fig:minimal-models}
\end{figure}

@@

In order for predicted bifurcations $p(\theta)\in\predictions$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. We expect this term to be some mean over the targets $\targets$ and predictions $\predictions(\theta)$. We choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target in cases where $|\predictions|\geq|\targets|$. This term only zero when each target is matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:supervised-term}
\end{align}
We can see from Figures \ref{fig:saddle-node} and \ref{fig:pitchfork} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$ in the non-degenerate case. Using these quantities we can define a positive semi-definite measure $\measure(z)$ of zero crossings in the determinant along the bifurcation curve; as shown in Figure \ref{fig:bifurcation-measure} this can be used as a proxy for the probability of a bifurcation at location $p(\theta)$. The total bifurcation measure $\Psi(\theta)$ is written
\begin{align}
    \Psi(\theta):=\frac{
        \int_{\rates(z)=0}\!
        \measure(z)\,\mathrm{d}z
    }{
        \int_{\rates(z)=0}\!\
        \mathrm{d}z
    }
    \quad\mathrm{where}\quad
    \measure(z):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{align}
Here we denote $\int_{\rates(z)=0}\mathrm{d}z$ as the sum of the line integrals in $z\in\Reals^{N+1}$ defined by the steady states $\rates(z)=0$. The measure $\measure(z)=1$ at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in $\frac{d}{ds}\Det\rightarrow0$ and hence $\measure(z)\rightarrow0$ for each turning point. As $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(z)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(z)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the observed parameter region $p$ is densely packed with bifurcations. The $\measure(z)$ is also invariant under certain classes of transformations on $\rates(z)$ which do not change bifurcation locations, proofs of which are provided in Appendix \ref{appendix:measure}. The measure is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:cost}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that in principle while individual bifurcations $p(\theta)$ depend smoothly on $\theta$ the total number of predictions $|\predictions|$ does not. The gradient $\partial_{\theta}|\predictions|=0$ everywhere and is an infinite impulse at the exact locations where the number of predictions changes.

>>>>

For clarity, we guide the reader through the methods with the following minimal models that explore the space of saddle-nodes $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchforks $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$. These minimal models are shown with targets $\targets$ in Figure \ref{fig:minimal-models}. These figures show that the determinant $\Det=0$ at the bifurcations and that its directional derivative along the bifurcation curve $\frac{d}{ds}\Det$ is finite and non-zero, at least in the non-degenerate case. This gives us our first hint of what should be optimised to increase the number of predictions $|\predictions(\theta)|$.

%   \centering
%   \setlength\unitlength{1cm}
%   {\phantomsubcaption\label{fig:1a}}
%   {\phantomsubcaption\label{fig:1b}}

% \textcolor{gray}{\rule{5cm}{3cm}}%
% \begin{picture}(0,0)
% \put(-4.8,2.6){\subref{fig:1a}}
% \end{picture}
% \textcolor{gray}{\rule{5cm}{3cm}}%
% \begin{picture}(0,0)
% \put(-4.8,2.6){\subref{fig:1b}}
% \end{picture}

\begin{figure}
\centering
\includegraphics[width=6cm]{saddle-node}
\includegraphics[width=6cm]{pitchfork}
\caption{Saddle-node model (left column) $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchfork model (right column) $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$ with $\theta=(5/2,-1),(1/2,-1)$ and targets $\targets=\{-1,1\},\{0\}$ respectively. Determinant in red. Lighter shades indicate the determinant crossing zero for unstable solutions}
\label{fig:minimal-models}
\end{figure}

@@

In order for predicted bifurcations $p(\theta)\in\predictions$ to match targets $p'\in\targets$ we need to evaluate an error term $|p(\theta)-p'|$. We expect this term to be some mean over the targets $\targets$ and predictions $\predictions(\theta)$. We choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target in cases where $|\predictions|\geq|\targets|$. This term only zero when each target is matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:supervised-term}
\end{align}
We can see from Figures \ref{fig:minimal-models} that predictions $p(\theta)$ can be identified by looking for points along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$ in the non-degenerate case. Using these quantities we can define a positive semi-definite measure $\measure(z)$ of zero crossings in the determinant along the bifurcation curve; as shown in Figure \ref{fig:bifurcation-measure} this can be used as a proxy for the probability of a bifurcation at location $p(\theta)$. The total bifurcation measure $\Psi(\theta)$ is written
\begin{align}
    \Psi(\theta):=\frac{
        \int_{\rates(z)=0}\!
        \measure(z)\,\mathrm{d}z
    }{
        \int_{\rates(z)=0}\!\
        \mathrm{d}z
    }
    \quad\mathrm{where}\quad
    \measure(z):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{align}
Here we denote $\int_{\rates(z)=0}\mathrm{d}z$ as the sum of the line integrals in $z\in\Reals^{N+1}$ defined by the steady states $\rates(z)=0$. The measure $\measure(z)=1$ at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in $\frac{d}{ds}\Det\rightarrow0$ and hence $\measure(z)\rightarrow0$ for each turning point. As $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(z)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(z)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the observed parameter region $p$ is densely packed with bifurcations. The $\measure(z)$ is also invariant under certain classes of transformations on $\rates(z)$ which do not change bifurcation locations, proofs of which are provided in Appendix \ref{appendix:measure}. The measure is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=
    \big(|\predictions|-|\targets|\big)\,\,\lambda \log\Psi(\theta)+
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \prod_{p(\theta)\in\predictions}|p(\theta)-p'|
    ^{\frac{1}{|\predictions|}}
    \label{eq:cost}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that in principle while individual bifurcations $p(\theta)$ depend smoothly on $\theta$ the total number of predictions $|\predictions|$ does not. The gradient $\partial_{\theta}|\predictions|=0$ everywhere and is an infinite impulse at the exact locations where the number of predictions changes.
